Associations of environmental and human factors with altered rodent populations in West Africa: A scoping review

### 2021-03-24 reports/manuscript.Rmd
mapping rodents

The earliest studies identified were from 1974 with the majority (62%) published since 2010. Most of the included studies were classified as rodent ecology research (54%) with the remainder studies on the risk of zoonoses (46%).

@@

@@

The trapping activity from the included studies was spread across 1,171 study sites. Twenty-four (24%) studies reported trapping only occurring at a single study site, 34 (33%) studies trapped at between two and five study sites the remaining 44 studies trapped at between six and 93 study sites. Trap sites were mapped to level one country administrative regions and density of sites per million square kilometers was calculated (see Figure \@ref(fig:density)). The areas with highest trapping densities included the capital cities of Niger, Sierra Leone, Senegal, Benin, Mali, Ghana and the Ivory Coast. Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore) and Eastern Sierra Leone were most intensively studied.

@@

  + Species trapped
  + Pathogens

>>>>

The earliest studies identified were from 1974 with the majority (62%) published since 2010. Most of the included studies were classified as rodent ecology research (54%) with the remainder studies on zoonotic diseases (46%).

@@

The trapping activity from the included studies was spread across 1,171 study sites. Twenty-four (24%) studies reported trapping only occurring at a single study site, 34 (33%) studies trapped at between two and five study sites the remaining 44 studies trapped at between six and 93 study sites. Trap sites were mapped to level one country administrative regions and density of sites per million square kilometers was calculated (see Figure \@ref(fig:density)). The areas with highest trapping densities included the capital cities of Niger, Sierra Leone, Senegal, Benin, Mali, Ghana and the Ivory Coast. Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore) and Eastern Sierra Leone were most intensively studied. Mapping to level two administrative regions is displayed in Supplementary figure \@ref(fig:supfig1).

@@

  + Calculating where possible for site/study effort

@@

Across all West African study sites there were 59,343 trapped small mammals (592 were trapped outside of West Africa), 2,830 (4.8%) trapped individuals were identified to order level (Rodentia), 5,329 (9%) were identified to genus level, the remaining 51,184 (86%) were identified to species level. Of the 131 distinct identified species trapped (see Supplementary table 1) the majority were from the order Rodentia (102) with Muridae (83) being the largest family of rodents, followed by Gliridae (6), Nesomyidae (4), Sciuridae (3), Ctenodactylidae (2) and Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (all 1). The remaining 29 species came from the orders of the Soricomorpha (25), Erinaceomorpha (3) and a single Afrosoricida.

More individuals were trapped in studies conducted to investigate the risk of zoonoses (35,892, 60%) than from ecology studies (23,451 40%). Animals trapped as part of ecological research were more often identified to species level (93%) than those trapped in zoonoses research (82%). The most commonly trapped genera of rodents were *Mastomys sp.* (22,665, 39.7%), *Rattus sp.* (7,718, 13.5%), *Mus sp.* (7,317, 12.8%), *Praomys sp.* (4,887, 8.6%) and *Arvicanthis sp.* (4,408, 7.72%). At the species level *Mastomys natalensis* (10,419, 20.2%) and *Mastomys erythroleucus* (7,184, 13.9%) were the most commonly trapped rodents, *Rattus rattus* (6,229, 12.1%), *Mus musculus* (6,030, 11.8%), *Arvicanthis niloticus* (4,101, 8%) and *Mastomys huberti* (3,106, 6%) made up the next most commonly trapped species. Presence and absence maps for these six most commonly trapped species are shown in Figure \@ref(fig:dist).

``` {r dist, fig.cap = "Presence and absence of the 6 most commonly trapped species in included studies, no adjustment made for trapping effort", echo = F, layout = "l-page"}
knitr::include_graphics(here::here("figures", "top_6.png"))
```

  + Pathogens

## Supplementary material

#### Supplementary figure 1

``` {r supfig1, fig.cap = "Density of trap sites per thousand km^2^", echo = F, layout = "l-page"}
knitr::include_graphics(here::here("figures", "static_site_density_2.png"))
```

#### Supplementary table 1

The number and percentage of trapped rodents identified to species level

``` {r sup_table, echo = F}
DT::datatable(readRDS(here::here("tables", "sup_table1.rds")))
```
